Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2020-05-04 manuscript/manuscript.tex
updated ms

\begin{align}
  G(t) = \frac{\sum_{i\ne j}^{t-1} \cov(\Delta p_i, \Delta p_j)}{\var(p_t - p_0)}
\end{align}
%
which estimates the effect of selection on allele frequency change between the
initial generation $0$ and some later generation $t$, which can be varied to
see how this quantity grows through time. Since \textcite{Barghi2019-qy}
experiment is sequenced every ten generations, in the numerator for the
covariance we use the allele frequency changes between adjacent timepoints,
which are ten generations apart. Consequently, this leads our measure $G(t)$ to
be strongly conservative, since the temporal covariances within each
ten-generation block are not directly observable, and thus are not included in
the numerator of $G(t)$. Still, we find a remarkably strong signal. Greater
than $20\%$ of total, genome-wide allele frequency change over 60 generations
is the result of selection (Figure \ref{fig:figure-1} B).

@@

  \caption{{\bf A}: The convergence \vb{correlations}, averaged across
    \textcite{Barghi2019-qy} replicate pairs, through time. Each line
    represents the convergence correlation $\mathrm{cor}(\Delta p_{s}, \Delta
    p_{s})$ from a starting reference generation $s$ to a later time $t$, which
    varies along the x-axis; each line corresponds to a row of the temporal
    convergence correlation matrix depicted to the right. We note that
    convergent correlation for the last timepoint is an outlier; we are unsure
    as to the cause of this, e.g. it does not appear to be driven by a single
    pair of replicates. {\bf B}: The convergence \vb{correlations} between
    individual pairs of replicates in the \textcite{Kelly2019-dc} data
    (\vb{note the confidence intervals are plotted, but are small on this
    y-axis scale; see note XXX}). {\bf C}:  The convergence \vb{correlations}
    between individual pairs of replicates in \parencite{Castro2019-uk} data,
    for the two selection lines  (LS1 and LS2) and the control (Ctrl); gray CIs
    are those using the complete dataset, blue CIs exclude chromosomes 5 and 10
    which harbor the two regions \textcite{Castro2019-uk} found to have signals
  of parallel selection between LS1 and LS2.}

@@

These results are often relatively insensitive to the number of loci underlying
the trait, suggesting that they may be capturing highly polygenic signals.
Indeed if only a small number of loci influence the trait, the $G(t)$
trajectories are typically much more stochastic, such that replicates do not
reliably generate temporal covariances (compare Figure \ref{fig:figure-1}B to
Supplementary Material Figure XXX).

Additionally, we explored other selection schemes. We find that the long term
dynamics of the covariances under truncation selection, which generates
substantial epistasis, are richer than we see under GSS and multiplicative
selection (Supplementary Material Figure XXX).  We also conducted simulations
of purifying selection alone (i.e. background selection) and find that this can
also generate positive temporal covariances (Figure XXX), thus it is unlikely
that all of the allele frequency change we see is the result of the novel
selection pressure the populations are exposed to, and some of this selection
may be ongoing purifying selection. Only in the case of the Longshanks
experiment, does the control line mean that we can quantify selection that is
due to the novel selection pressure.

@@

Our analyses reveal that a sizable proportion of allele frequency change in
these populations is due to the (indirect) action of selection. Capitalizing on
replicated designs, we characterized the extent to which convergent selection
pressures lead to parallel changes in allele frequencies across replicate
populations, and found that a reasonable proportion of the response is shared
across short timescales. These likely represent substantial under-estimates of
the contribution of linked selection because the studies we have reanalyzed do
not sequence the population each generation, preventing us from including the
effects of stronger correlations between adjacent generations.  Furthermore,
our estimation methods are intentionally conservative, for example they exclude
the contribution of selection that does not persist across generations and
selection that reverses sign; thus they can be seen as a strong lower bound of
the effects of selection.

@@

%By using an exponential fitness function (corresponding to
%multiplicative selection across sites), \vb{we ensure there is no build up of
%LD beyond what is initially present due to selection and epistasis}
%\parencite{Turelli1990-kd,Bulmer1980-zo}; this serves as the simplest
%directional selection model of a trait to understand the effects of genetic
%architecture on the statistics we have used in the paper.

@@

\subsubsection{Sampling in Temporal Blocks}

>>>>

\begin{align}
  \vb{G(t) = \frac{ \sum_{i=0}^{t-1} \sum_{j \ne i}^{t-1} \cov(\Delta p_i, \Delta p_j)}{\var(p_t - p_0)}}
\end{align}
%
which estimates the effect of selection on allele frequency change between the
initial generation $0$ and some later generation $t$, which can be varied to
see how this quantity grows through time. Since \textcite{Barghi2019-qy}
experiment is sequenced every ten generations, in the numerator for the
covariance we use the allele frequency changes between adjacent timepoints,
which are ten generations apart; \vb{thus the strong covariances between
  adjacent generations are not observable and do not contribute to the
  numerator of $G(t)$}. \vb{Additionally, selection also inflates the variance
  in allele frequency change per generation; however, this effect cannot be
  easily distinguished from drift.} \vb{For both these reasons, our measure
  $G(t)$ is strongly conservative} (\vb{we demonstrate this through simulations
in Supplementary Material Section \ref{supp:tempblocks}}). Still, we find a
remarkably strong signal. Greater than $20\%$ of total, genome-wide allele
frequency change over 60 generations is the result of selection (Figure
\ref{fig:figure-1} B).

@@

  \caption{{\bf A}: The convergence \vb{correlations}, averaged across
    \textcite{Barghi2019-qy} replicate pairs, through time. Each line
    represents the convergence correlation $\mathrm{cor}(\Delta p_{s}, \Delta
    p_{s})$ from a starting reference generation $s$ to a later time $t$, which
    varies along the x-axis; each line corresponds to a row of the temporal
    convergence correlation matrix depicted to the right \vb{(where the
      diagonal elements represent the convergence correlations between the same
      timepoints across replicate populations)}. We note that convergent
      correlation for the last timepoint is an outlier; we are unsure as to the
      cause of this, e.g. it does not appear to be driven by a single pair of
      replicates. {\bf B}: The convergence \vb{correlations} between individual
      pairs of replicates in the \textcite{Kelly2019-dc} data (\vb{note the
      confidence intervals are plotted, but are small on this y-axis scale}).
      {\bf C}:  The convergence \vb{correlations} between individual pairs of
      replicates in \parencite{Castro2019-uk} data, for the two selection lines
      (LS1 and LS2) and the control (Ctrl); gray CIs are those using the
      complete dataset, blue CIs exclude chromosomes 5 and 10 which harbor the
      two regions \textcite{Castro2019-uk} found to have signals of parallel
      selection between LS1 and LS2. \vb{Through simulations, we have found
        that the differences in convergence correlation confidence interval
        widths between these \emph{Drosophila} studies and the Longshanks study
        are due to the differing population sizes.}}

@@

These results are often relatively insensitive to the number of loci underlying
the trait, suggesting that they may be capturing highly polygenic signals.
Indeed if only a small number of loci influence the trait, the $G(t)$
trajectories are typically much more stochastic, such that replicates do not
reliably generate temporal covariances (compare Figure \ref{fig:figure-1}B to
Supplementary Material Figure \ref{suppfig:sim-expfit-covs}).

Additionally, we explored other selection schemes. We find that the long term
dynamics of the covariances under truncation selection, which generates
substantial epistasis, are richer than we see under GSS and multiplicative
selection (Supplementary Material Figure \ref{suppfig:supp-trunc}).  We also
conducted simulations of purifying selection alone (i.e. background selection)
and find that this can also generate positive temporal covariances
(Supplementary Material Figure \ref{suppfig:supp-bgs-covs}), thus it is
unlikely that all of the allele frequency change we see is the result of the
novel selection pressure the populations are exposed to, and some of this
selection may be ongoing purifying selection. Only in the case of the
Longshanks experiment, does the control line mean that we can quantify
selection that is due to the novel selection pressure.

@@

Our analyses reveal that a sizable proportion of allele frequency change in
these populations is due to the (indirect) action of selection. Capitalizing on
replicated designs, we characterized the extent to which convergent selection
pressures lead to parallel changes in allele frequencies across replicate
populations, and found that a reasonable proportion of the response is shared
across short timescales. These likely represent substantial under-estimates of
the contribution of linked selection because the studies we have reanalyzed do
not sequence the population each generation, preventing us from including the
effects of stronger correlations between adjacent generations.  Furthermore,
our estimation methods are intentionally conservative, for example they exclude
the contribution of selection that does not persist across generations and
selection that reverses sign; thus they can be seen as a strong lower bound of
the effects of selection, \vb{which we have confirmed through forward-in-time
simulations}.

@@

@@

\subsubsection{Sampling in Temporal Blocks}
\label{supp:tempblocks}
